What is an Ancestral Recombination Graph?

### 2022-02-10 paper.tex
Small tweaks to text

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of species \citep{rannala2003genetics}, non (or
rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}. The tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these trees [cite?], and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the
trees~\citep{tajima1983evolutionary,tavare1984line,fu1995statistical}
such as allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical genetics.

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2021unified}
and data representation~\citep{kelleher2016efficient}
have raised the genuine prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit.
Applications using these inferred ancestries as input have
begun to appear~\citep{osmond2021estimating} and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies,
leading to basic errors in statements about what different inference methods
produce as output (see the XXX section below), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a users perspective) using multiple inference methods in an analysis.

@@

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2); % D
	\draw[color=gray, dashed] (7.5, 1.5) -- (11, 1.5); % E and F
	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4); % G and H
	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1); % I and J
	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);	% P and Q
	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5); % K
	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4); % L
	\draw[color=gray, dashed] (7.5, 6) -- (11, 6); % M and N
	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);	% R
	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2); % O
\end{tikzpicture}
\end{tabular}
}
\caption{(Top) A realisation of the graph traversed by Hudson's algorithm started from a sample of three continuous chromosomes of length $L$ at time $t = 0$, and propagated until time $T$.
The MRCA on the genetic interval $[x, z)$ is reached at node L, while that on $[0, x)$ is reached at node O, which is a root as it carries only fully coalesced ancestral segments.
The non-ancestral segment $[x, y)$ above node K contributes to the rate of effective recombinations because it is trapped between ancestral segments.
The two columns titled ``Extant material" are the number of extant lineages and the total length of genetic material on which an effective recombination can take place during each time interval.
(Bottom) A corresponding realisation of a big ARG, which augments Hudson's algorithm by tracking nonancestral lineages.
The result is a simpler state space and dynamics, at the cost of extra nodes and edges, highlighted in red, which do not affect the local tree at any site.}
\label{hudson_vs_bigARG}
\end{figure}

Hudson's algorithm operates by tracking the state of a set of ancestral
lineages as we go backwards in time.
% jk-note unclear here about the best choice of notation. We could have
% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
% stick with the original tuple notation for now.
%
% Each lineage consists of a set of
% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
% such that for each $1 \leq j \leq \left| l \right|$,
% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
% $\mathbf{a}_j$ is an integer
% tracking the number of samples the lineage is ancestral to over the interval.
% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
% covered by the lineage, respectively.
Each lineage consists of a list of
disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples to which the lineage is ancestral over that interval.
(We also usually track the tree node associated with each segment, but
that is not important for our purposes here so we omit it.)
If we have $n$ samples and genome of length $m$, the process begins with $n$ lineages
of the form $[(0, m, 1)]$. The process then works backwards in time from
the present day as a series of random common ancestor or recombination events.
Recombination events occur at a rate determined by the amount of ancestral material and
the way in which it is distributed along a chromosome.
Let $L$ be the set of ancestors at a given time $t$. Recombination events
happen at rate $\rho \nu / (m - 1)$ where
\[
\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
\]
is the number of available `links' that may be broken. Thus, the rate of
recombination is determined by the left- and right-most extent of the
ancestral material carried by each lineage. At a recombination
event we choose one of these links uniformly and break it. Given a lineage
$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
$x_1$ and $x_2$ such that FILL IN DETAILS

@@

The state-space of this process is much simpler than Hudson's algorithm, which
greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the ARG back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the complete ancestry of a sample
in Hudson's algorithm is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the vast majority of the events that occur in the ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''

>>>>

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of species \citep{rannala2003genetics}, non (or
rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}.
A genealogical tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these trees [cite?], and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the
trees~\citep{tajima1983evolutionary,tavare1984line,fu1995statistical}
such as allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical
genetics.

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2021unified}
and data representation~\citep{kelleher2016efficient}
have raised the genuine prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have
begun to appear~\citep{osmond2021estimating} and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies,
leading to basic errors in statements about what different inference methods
produce as output (see the XXX section below), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a users perspective) using multiple inference methods in an analysis.

@@

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2); % D
	\draw[color=gray, dashed] (7.5, 1.5) -- (11, 1.5); % E and F
	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4); % G and H
	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1); % I and J
	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);	% P and Q
	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5); % K
	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4); % L
	\draw[color=gray, dashed] (7.5, 6) -- (11, 6); % M and N
	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);	% R
	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2); % O
\end{tikzpicture}
\end{tabular}
}
\caption{(A)
A realisation of the graph traversed by Hudson's algorithm started from a
sample of three continuous chromosomes of length $L$ at time $t = 0$, and
propagated until time $T$. The MRCA on the genetic interval $[x, z)$ is reached
at node L, while that on $[0, x)$ is reached at node O.
The non-ancestral segment $[x, y)$ above
node K contributes to the rate of effective recombinations because it
is trapped between ancestral segments. The two columns titled ``Extant
material" are the number of extant lineages and the total length of genetic
material on which an effective recombination can take place during each time
interval.
(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
by tracking nonancestral lineages. The result is a simpler state space and
dynamics, at the cost of extra nodes and edges, highlighted in red, which do
not affect the local tree at any site.}
\label{hudson_vs_bigARG}
\end{figure}

Hudson's algorithm operates by tracking the state of a set of ancestral
lineages as we go backwards in time.
% jk-note unclear here about the best choice of notation. We could have
% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
% stick with the original tuple notation for now.
%
% Each lineage consists of a set of
% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
% such that for each $1 \leq j \leq \left| l \right|$,
% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
% $\mathbf{a}_j$ is an integer
% tracking the number of samples the lineage is ancestral to over the interval.
% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
% covered by the lineage, respectively.
Each lineage consists of a list of
disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples to which the lineage is ancestral over that interval.
(We also usually track the tree node associated with each segment, but
that is not important for our purposes here so we omit it.)
If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
of the form $\{(0, m, 1)\}$. The process then works backwards in time from
the present day as a series of random common ancestor or recombination events.
Recombination events occur at a rate determined by the amount of ancestral material and
the way in which it is distributed along a chromosome.
Let $L$ be the set of ancestors at a given time $t$. Recombination events
happen at rate $\rho \nu / (m - 1)$ where
\[
\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
\]
is the number of available `links' that may be broken. Thus, the rate of
recombination is determined by the left- and right-most extent of the
ancestral material carried by each lineage. At a recombination
event we choose one of these links uniformly and break it. Given a lineage
$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
$x_1$ and $x_2$ such that FILL IN DETAILS

@@

The state-space of this process is much simpler than Hudson's algorithm, which
greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the ARG back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the complete ancestry of a sample
in Hudson's algorithm is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the majority of the events that occur in the ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events that occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''
